---
title: "PredictDB weight distribution"
author: "Haky Im"
date: "2023-02-28"
categories: [analysis]
format:
  html:
    code-fold: true
    code-summary: "Show the code"
editor_options: 
  chunk_output_type: console
---

Goal: get effect size distribution of omic traits

```{r preliminary definitions}
suppressMessages(library(tidyverse))
suppressMessages(library(glue))
PRE = "/Users/haekyungim/Library/CloudStorage/Box-Box/LargeFiles/imlab-data/data-Github/web-data"
##PRE="/Users/margaretperry/Library/CloudStorage/Box-Box/imlab-data/data-Github/web-data "
##PRE="/Users/temi/Library/CloudStorage/Box-Box/imlab-data/data-Github/web-data"
## COPY THE DATE AND SLUG fields FROM THE HEADER
SLUG="predictdb-weight-distribution" ## copy the slug from the header
bDATE='2023-02-28' ## copy the date from the blog's header here
DATA = glue("{PRE}/{bDATE}-{SLUG}")
if(!file.exists(DATA)) system(glue::glue("mkdir {DATA}"))
WORK=DATA
USERNAME="haekyungim"
## move data to DATA
#tempodata=("~/Downloads/tempo/gwas_catalog_v1.0.2-associations_e105_r2022-04-07.tsv")
#system(glue::glue("cp {tempodata} {DATA}/"))
## system(glue("open {DATA}")) ## this will open the folder 
```

-   [ ] download mashr gtex v8 prediction models for various

```{r, eval=FALSE}
if(F)
{
  setwd(DATA)
  system(glue("wget https://zenodo.org/record/3518299/files/mashr_eqtl.tar?download=1"))
  system(glue("tar xvf mashr_eqtl.tar"))
  system(glue("cd eqtl"))
  system("rm *.txt.gz")
}
```

-   [ ] plot distribution, show summary

```{r}
library("RSQLite")
sqlite <- dbDriver("SQLite")

myshowdist = function(dbname,titulo="n=")
{
print(dbname)
db = dbConnect(sqlite,dbname)
weights = dbGetQuery(db, "select * from weights")
absweivec = abs(weights$weight)
qlines = quantile(absweivec,c(0.2,0.5)) %>% round(2)
quantile(absweivec,(0:10)/10) %>% round(2) %>% print()
pp <- weights %>% mutate(abswei=abs(weight)) %>% ggplot(aes(abswei))+geom_histogram() + geom_vline(xintercept=qlines) +ggtitle(glue("{titulo}; 20%={qlines[1]}, median={qlines[2]}"))
print(pp)
dbDisconnect(db)
}
```

-   [ ] read weights adipose with 491 samples

```{r}
dbname <- glue("{DATA}/eqtl/mashr_Adipose_Subcutaneous.db") ## add full path if db file not in current directory
myshowdist(dbname,titulo="Adipose expr n=491")
```

-   [ ] read weights brain substantia nigra with 101 samples

```{r}
dbname <- glue("{DATA}/eqtl/mashr_Brain_Substantia_nigra.db") ## add full path if db file not in current directory
myshowdist(dbname,titulo="Brain SN expr - n=101")
```

-   [ ] read protein prediction weights

```{r}
## can download from https://uchicago.box.com/shared/static/m3zsxy3oy8kn5gkktkuo8lvui269hxi5.db
print("AA")
dbname <- glue::glue("/Users/{USERNAME}/Library/CloudStorage/Box-Box/LargeFiles/imlab-data/Within-Lab-Sharing/Sabrina-Data/ARIC/ARIC_AA_hg38.db")
myshowdist(dbname,titulo="protein ARIC AA n=1,871")
```

```{r}
## can download from https://uchicago.box.com/shared/static/m3zsxy3oy8kn5gkktkuo8lvui269hxi5.db
print("EUR")
dbname <- glue::glue("/Users/{USERNAME}/Library/CloudStorage/Box-Box/LargeFiles/imlab-data/Within-Lab-Sharing/Sabrina-Data/ARIC/ARIC_EA_hg38.db")
myshowdist(dbname,titulo="protein ARIC EA n=7,213")
```

-   [ ] read brainxcan weights (use elastic net to be closer to QTL effect size)

```{r}
##install.packages("arrow")
tempo = arrow::read_parquet(glue::glue("{DATA}/brainxcan-gw_lasso_beta.parquet"))
tempo = as.matrix(tempo %>% select(starts_with("IDP")))
kk = abs(tempo[tempo!=0])
print(summary(kk))
print(quantile(kk,(0:10)/10))
print(quantile(kk,c(0.2,0.5)))
```

## get protein predictors with FDR < 0.05
just need to count the number of prediction models
```{r}
print("Protein ARIC models AA")
dbname <- glue::glue("/Users/{USERNAME}/Library/CloudStorage/Box-Box/LargeFiles/imlab-data/Within-Lab-Sharing/Sabrina-Data/ARIC/ARIC_AA_hg38.db")
print(dbname)
db = dbConnect(sqlite,dbname)
##weights = dbGetQuery(db, "select * from weights")
extra = dbGetQuery(db,"select * from extra")
nrow(extra)
dbDisconnect(db)

print("Protein ARIC models EUR")
dbname <- glue::glue("/Users/{USERNAME}/Library/CloudStorage/Box-Box/LargeFiles/imlab-data/Within-Lab-Sharing/Sabrina-Data/ARIC/ARIC_EA_hg38.db")
print(dbname)
db = dbConnect(sqlite,dbname)
##weights = dbGetQuery(db, "select * from weights")
extra = dbGetQuery(db,"select * from extra")
nrow(extra)
dbDisconnect(db)

```

```{r}
dbname <- glue("{DATA}/en_Adipose_Subcutaneous.db") ## add full path if db file not in current directory
print("Adipose models")
print(dbname)
db = dbConnect(sqlite,dbname)
##weights = dbGetQuery(db, "select * from weights")
extra = dbGetQuery(db,"select * from extra")
nrow(extra)
dbDisconnect(db)


dbname <- glue("{DATA}/en_Brain_Substantia_nigra.db") ## add full path if db file not in current directory
print("Brain Substantia nigra models")
print(dbname)
db = dbConnect(sqlite,dbname)
##weights = dbGetQuery(db, "select * from weights")
extra = dbGetQuery(db,"select * from extra")
nrow(extra)
dbDisconnect(db)
```

