---
title: How to calculate Z-score, P-value, Chi2 stat from GWAS
author: Haky Im
date: '2022-04-13'
---


::: {.callout-tip}
## Summary
Given one of the statistics in a GWAS (Z-score, P-value, or chi2), calculate the others.
:::


## Calculate Zscore, p-value, Chi2 statistics using GWAS summary statistics

$$Y = \beta \cdot X + \epsilon$$
GWAS summary statistics will contain an estimate of the regression coefficient $\hat\beta$ and its standard error $\text{se}(\hat\beta)$ for each SNP in the GWAS.

We distinguish the true $\beta$ from the estimated value $\hat\beta$ using a hat.

## Z-score

$$ Z = \frac{\hat\beta}{\text{se}(\hat\beta)}$$
$$Z \approx N(0,1) ~~~~~~ \text{as } n \rightarrow \infty$$


## Z to Chi2 statistic

$$Z^2 \sim \chi^2_\text{df=1}$$

Under the null, the squared of the Z-score follows a $\chi^2$ distribution with 1 degree of freedom.

## Z to P-value

$$P = \text{pnorm}(-|Z|) * 2$$

## P to Z

From the p-value, we can calculate the magnitude of the Z-score but the sign is lost. So the Z-score has more information than the p-value.

$$|Z| = |\text{qnorm}(P/2)|$$

## P to Chi2

$$ \chi^2 =  \text{qnorm}(P/2)^2 $$

-------

## Simulations 

Let's simulate zscore vector under the null hypothesis.

```{r}
nsim = 5000
set.seed(2021021001)
zvec = rnorm(5000, mean=0, sd=1)
```

Calculate the p-value (probability that a normal r.v. will be as large or larger in magnitude than the |zscore| 
```{r}
pvec = pnorm(-abs(zvec)) * 2 ## two-tailed
## check pvec is uniformly distributed
hist(pvec,main="is this uniformly distributed?")
## remember that if square the a normal r.v. you get chi2 r.v. with one degree of freedom
z2vec = zvec^2
## compare with chi2 rv. with 1 degree of freedom by simulating chi2,1 and qqplot
chi2vec = rchisq(nsim,df=1)
qqplot(chi2vec,z2vec,main="compare of zscore^2 and a chi2 r.v."); abline(0,1)
```

```{r}
## test whether the distributions of z2vec and chi2vec are different using the Kologorov Smirnov test
ks.test(z2vec,chi2vec)
```


```{r}
## for reference, let's compare two chi2,1 r.v.'s qqplot
chi2vec_2 = rchisq(nsim,df=1)
qqplot(chi2vec,chi2vec_2,main="qqplot of two chi2,1 r.v."); abline(0,1)
ks.test(chi2vec,chi2vec_2)
```

> Sometimes you get the p-value instead of the zscore, you can generate chi2 by inverting the relationship.

```{r}
chi2vec_from_p = qnorm(pvec / 2)^2
qqplot(chi2vec,chi2vec_from_p,main="chi2 from pvalue vs chi2 r.v."); abline(0,1)
```
